Data Preparation

### 2021-07-12 paper.Rmd
static impact/disturbance levels included to method section (next: sliding window by Fabio)

This paper is heavily informed by the work of [@strohmeier_crowdsourced_2021].

@@

@@

@@

>>>>

@@

The global daily flight dataset by Opensky Network identifies a flight based on a series of received messages of 15 minutes or more [@strohmeier_crowdsourced_2021].
As the sensor network does not cover the whole globe/airspace, a consistency check is made for flights leaving the coverage area and entering again (e.g. oceanic area).
A heuristic algorithm for the identification of destination airports is applied for descending trajectories. 
While there are uncertainties in the data based on these heuristics, the dataset ensures detection of flights within certain geographic regions even without a positive departure and/or destination aerodrome identification.

@@

> TODO: develop table with months - file size - number of flights

@@

```{r}
# load analytic dataset
us <- ds %>% 
  filter(ADEP_REG == "US" | ADES_REG == "US") %>%
  group_by(YR = year(DATE), DATE) %>% 
  summarize(N = n(), .groups = "drop")

# static thresholds, x-th percentile rounded to lower integer
pcts   <- c(0.8, 0.6, 0.3, 0.1)
us_ref <- us %>% filter(YR == 2019) %>%
  pull(N) %>% quantile(probs = pcts) %>% floor()

us_ref2 <- us %>% filter(YR == 2019, DATE >= lubridate::ymd("2019-07-01")) %>%
  pull(N) %>% quantile(probs = pcts) %>% floor()

max_N <- max(us$N)
lvls  <- c("First", "Second", "Third", "Fourth", "Fifth")
us <- us %>% 
  mutate( THRES  = cut(N, breaks = c(max_N, us_ref, 0),  labels = lvls)
         ,THRES2 = cut(N, breaks = c(max_N, us_ref2, 0), labels = lvls)
         )
```

(ref:fig-us-timeline) Resilience concept applied to observed US air traffic development

```{r fig-us-timeline, fig.cap="(ref:fig-us-timeline)"}
us %>%
  ggplot(aes(x = DATE)) +
  geom_hline(yintercept = us_ref2, linetype = "dotdash") +
  geom_point(aes(y = N, color = THRES2), alpha = 0.3) +
  geom_line( aes(y = N, colour = THRES2, group = 1)) +
  theme_minimal()
```

> TODO - further develop & include sliding window approach proposed by Fabio

### 2021-07-12 x0_data-preparation.Rmd
static impact/disturbance levels included to method section (next: sliding window by Fabio)

```{r, eval=TRUE}
#quick peek

@@

>>>>

```{r, eval=TRUE, message=FALSE}
#quick peek

@@

```{r, dpi=480}
#THRESHOLD AT ROLLING QUANTILES PER YEAR
#Please install the "roll" package.
#install.packages("roll")
library(roll)

ds1 <- full_dataset %>% group_by(YR = year(DATE), DATE) %>% summarize(N = n(), .groups = "drop") %>% mutate(THRESHOLD = quantile(N, probs = 0.7))
ds1

ds1 %>% ggplot(aes(x = DATE)) +
  geom_line(aes(y = THRESHOLD)) +
  geom_point(aes(y = N, color = N > THRESHOLD))
```

```{r}
#CHECK WEEKLY DISTRIBUTION
ds2 <- ds1 %>% 
  mutate( WEEK_DAY = wday(DATE, label = TRUE)
         ,W_END    = WEEK_DAY %in% c("Sat", "Sun"))
uplim <- .95
lowlim <- .20

ds2 <- ds2 %>% mutate(MED = roll_quantile(N, width = 7, p = .50, min_obs = 7),
                      UP_THR = roll_quantile(N, width = 180, p = uplim, min_obs = 7),
                      LOW_THR = roll_quantile(N, width = 180, p = lowlim, min_obs = 7))
ds2

ds2 %>% ggplot(aes(x = DATE)) +
  geom_col(aes(y = MED, fill = !(MED < LOW_THR | MED > UP_THR))) +
  geom_line(aes(y = UP_THR), color = "blue") +
  geom_line(aes(y = LOW_THR), color = "red") +
  scale_x_date(limits = c(as.Date("2019-10-01"), as.Date("2020-12-01")), date_breaks = "30 days", date_labels = "%b\n%y") +
  labs(y = "7-day rolling median", x = NULL, fill = "Normal operation?",
       title = "What is normal?",
       subtitle = "Red line - Lower threshold; Blue line - upper threshold",
       caption = "Lower threshold set to 20th percentile of the last 180 days;\nUpper threshold set to 95th percentile of the last 180 days"
)
```

```
# breadcrumb ------------- need to check what this is for --------------------
#ds2 %>% ggplot() +
  geom_freqpoly(aes(x = N, color = W_END))
```

```{r}
us <- full_dataset %>% 
  filter(ADEP_REG == "US" | ADES_REG == "US") %>%
  group_by(YR = year(DATE), DATE) %>% 
  summarize(N = n(), .groups = "drop")

# static thresholds, x-th percentile rounded to lower integer
pcts   <- c(0.8, 0.6, 0.3, 0.1)
us_ref <- us %>% filter(YR == 2019) %>%
  pull(N) %>% quantile(probs = pcts) %>% floor()

us_ref2 <- us %>% filter(YR == 2019, DATE >= lubridate::ymd("2019-07-01")) %>%
  pull(N) %>% quantile(probs = pcts) %>% floor()

max_N <- max(us$N)
lvls  <- c("First", "Second", "Third", "Fourth", "Fifth")
us <- us %>% 
  mutate( THRES  = cut(N, breaks = c(max_N, us_ref, 0),  labels = lvls)
         ,THRES2 = cut(N, breaks = c(max_N, us_ref2, 0), labels = lvls)
         )
```
```{r}
us %>%
  ggplot(aes(x = DATE)) +
 # geom_line(aes(y = THRES)) +
  geom_point(aes(y = N, color = THRES2)) +
  geom_line(aes(y = N, colour = THRES2, group = 1)) +
  geom_hline(yintercept = c(max_N, us_ref2, 0))
```
